Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-08-29 paper/intro.tex
some fairly small edits and rewrites. adding some references

\subsection{Gyrochronology}
It is well established that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence (MS) stars allows them to be dated via
their rotation periods in a practice known as gyrochronology, which is useful
since the ages of MS stars are extremely difficult to measure via the
traditional age-dating method of isochrone placement.
It is also well established that stars of the same age but different masses
have different rotation periods \racomment{(citation)}, thought to be caused
by the deeper convective zones, and therefore stronger magnetic dynamos (and
more efficient magnetic braking) in lower-mass stars.
However, an underlying assumption behind many empirical gyrochronology
relations is that the relationships between rotation period and photometric
color\footnote{As the directly observable quantity, color is usually used as a
mass-proxy, and empirical gyrochronology relations are usually calibrated in
color, rather than mass or effective temperature.
}, and rotation period and age are {\it separable}, meaning that the
period-color relation is the same at all ages and the period-age relation is
the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
meibom2011, angus2015, angus2019}.
It was recently shown that old field stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \citet{skumanich1972} spin-down law is required
to reproduce the data \citep{vansaders2016, vansaders2018}.
An even more recent analysis of the 1.1 Gyr open cluster, NGC 6811, provides
further indication that the exponent of the period-age relation is
mass and time-dependent.
This cluster has a flattened relationship between rotation period and color:
the G dwarfs rotate at the same rate as the K dwarfs \citep{curtis2019}.
% Since a separable period-age and period-color relation provided an adequate
% fit for the young stars, it was naturally assumed to apply to old stars too.

In this paper, we tested the \citet{angus2019} gyrochronology relation, which
was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it the ideal cluster for calibrating the period-color
relation of stars at 650 Myrs.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
This relation accurately describes the rotation periods of F and G stars up to
around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
rotation periods), but over-predicts the rotation periods of K dwarfs in the
1.1 Gyr NGC 6811 cluster.
No reliable age estimates exist for K dwarfs with rotation periods older than
1.1 Gyr, so the rotational evolution of these stars beyond this age is a
mystery.
% In this paper, we shed light on the rotational evolution of old K dwarfs using
% population-based dynamical ages.
% Unfortunately, open clusters with rotation period measurements are mostly
% young -- currently the oldest is 2.5 Gyr (NGC 6819).
% Without rotation periods for precisely dated old stars, it is extremely
% difficult to calibrate the relationship between rotation period and color at
% old ages.
For this reason, we used a population-based stellar age indicator, velocity
dispersion, to investigate the period-color relations at old ages.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the center of the galaxy with a low out-of-plane, or vertical, velocity ($W$,
or $v_z$).
% because the MW's gas disk is thin and its gas moves in circular
% orbits.
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds, spiral arms and the galactic bar are thought to
play an important role \citep[see][for a review of secular evolution in the
MW]{sellwood2014}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
% Stellar velocities have a history of being used as an age proxy, with several
% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
% west2011}.
Since the age-velocity dispersion relations (AVRs) are themselves still
actively being calibrated, it is difficult to directly compare gyrochronal
ages with kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be a monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.

% Vertical {\it actions} are better age indicators than velocities, because
% actions are calculated by integrating angular momentum over the Milky Way's
% potential, and are therefore position invariant -- \eg\ a star will have the
% same action at periapsis and apoapsis.
% In contrast, orbital velocities are different at periapsis and apoapsis -- so
% in this sense, {\it actions} are the natural quantities to use as age proxies.
Vertical action is a better age indicator than vertical velocity
\citep[although still only a weak age indicator for individual
stars,][]{beane2018}, however both vertical action ($J_z$) and vertical velocity
(\vz/W) can only be calculated with full 6-dimensional position and velocity
information.
Unfortunately, most stars with measured rotation periods do not have radial
velocity (RV) measurements because they are relatively faint \kepler\ targets
($\sim$11th-18th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close (although not perfect -- see section \ref{sec:results})
approximation to \vz.

>>>>

\subsection{Gyrochronology}
It is well established that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence (MS) stars allows them to be dated using
their rotation periods, via gyrochronology \citep[\eg][]{kawaler1989,
pinsonneault1989, barnes2003, barnes2007, barnes2010, meibom2011, meibom2015},
which is particularly useful since the ages of MS stars are difficult to
measure via isochrone placement.
It is also well established that stars of the same age but different masses
have different rotation periods \citep[\eg][]{kraft1967, matt2012}, thought to
be caused by the deeper convective zones, and therefore stronger magnetic
dynamos (and more efficient magnetic braking) in lower-mass stars.
However, an underlying assumption behind many empirical gyrochronology
relations is that the relationships between rotation period and photometric
color\footnote{As the directly observable quantity, color is often used as a
mass proxy, and empirical gyrochronology relations are usually calibrated in
color, rather than mass or effective temperature.},
and rotation period and age are {\it separable}, meaning that the
period-color relation is the same at all ages and the period-age relation is
the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
meibom2011, angus2015, angus2019}.
Not all gyrochronology models follow this form, particularly those motivated
based on stellar structure and evolution theory or with a Rossby number,
rather than rotation period dependence \citep[\eg][]{barnes2010, matt2012,
vansaders2016]}.
It was recently shown that old field stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \prot\ $\propto t^{\frac{1}{2}}$ spin-down law
\citep{skumanich1972} is required to reproduce the data \citep{vansaders2016,
vansaders2018}.
An even more recent analysis of the 1.1 Gyr open cluster, NGC 6811, provides
further indication that the exponent of the period-age relation is
mass and time-dependent \citep{curtis2019}.
This cluster has a flattened relationship between rotation period and color
and the G dwarfs rotate at the same rate as the K dwarfs.
% Since a separable period-age and period-color relation provided an adequate
% fit for the young stars, it was naturally assumed to apply to old stars too.

In this paper, we tested the \citet{angus2019} gyrochronology relation, a
separable relation, calibrated using the period-color relation of Praesepe (in
\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it the ideal cluster for calibrating the period-color
relation of stars at 650 Myrs.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
This relation accurately describes the rotation periods of F and G stars up to
around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
rotation periods), but over-predicts the rotation periods of K dwarfs in the
1.1 Gyr NGC 6811 cluster.
Very few reliable age estimates exist for K dwarfs with rotation periods older
than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
evolution of middle-aged K dwarfs is a mystery\footnote{The oldest M dwarfs
with reliable ages and rotation periods are members of Praesepe ($\sim$ 650
Myrs), so even less is known about their rotational evolution.}.
% In this paper, we shed light on the rotational evolution of old K dwarfs using
% population-based dynamical ages.
% Unfortunately, open clusters with rotation period measurements are mostly
% young -- currently the oldest is 2.5 Gyr (NGC 6819).
% Without rotation periods for precisely dated old stars, it is extremely
% difficult to calibrate the relationship between rotation period and color at
% old ages.
For this reason, we used a population-based stellar age indicator, velocity
dispersion, to investigate the period-color relations at old ages in the
field.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the center of the galaxy with a low out-of-plane, or vertical, velocity ($W$,
or $v_z$), just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
% because the MW's gas disk is thin and its gas moves in circular
% orbits.
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time \citep[\eg][]{nordstrom2004, holmberg2007,
holmberg2009, aumer2009, casagrande2011}.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds, spiral arms and the galactic bar are thought to
play an important role \citep[see][for a review of secular evolution in the
MW]{sellwood2014}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
% Stellar velocities have a history of being used as an age proxy, with several
% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
% west2011}.
The age-velocity dispersion relations (AVRs) are still actively being
calibrated, so it is difficult to directly compare gyrochronal ages with
kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be a monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.

% Vertical {\it actions} are better age indicators than velocities, because
% actions are calculated by integrating angular momentum over the Milky Way's
% potential, and are therefore position invariant -- \eg\ a star will have the
% same action at periapsis and apoapsis.
% In contrast, orbital velocities are different at periapsis and apoapsis -- so
% in this sense, {\it actions} are the natural quantities to use as age proxies.
Vertical action is a better age indicator than vertical velocity
\citep{ting2019}, although still only a weak age indicator for individual
stars \citep{beane2018}, however both vertical action ($J_z$) and vertical
velocity (\vz/W) can only be calculated with full 6-dimensional position and
velocity information.
Unfortunately, most stars with measured rotation periods do not have radial
velocity (RV) measurements because they are relatively faint \kepler\ targets
($\sim$11th-18th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close (although imperfect -- see section \ref{sec:results})
approximation to \vz.

### 2019-08-29 paper/method.tex
some fairly small edits and rewrites. adding some references

We used the publicly available \kepler-\gaia\ DR2 crossmatched
catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ catalog of
stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
DR2 catalog.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
\citep{green2018}, and {\tt astropy} \citep{astropy2013}.
We estimated effective temperatures from dereddened \Gaia\ \gcolor\ color,
using an 8th-order polynomial relation calibrated using .... stars
\racomment{ask Jason for details}.
\begin{equation}
    \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
    301C^6 - 26C^7,
% 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
%           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
%           301.13205924630165, -25.923997443169355]
\label{eq:curtis}
\end{equation}
where C is \gaia\ \gcolor.

% \begin{figure}
%   \caption{
% A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
%     extinction-corrected magnitudes, colored by rotation period.
% We excluded photometric binaries and subgiants from our analysis by removing
% stars above the two dashed lines.
% % The rotation periods of binaries and subgiants do not follow a Skumanich-like
% % braking law.
% % The rotation period gradient across the main sequence is visible by eye in
% %     this figure: young, rapidly rotating stars are located below the old,
% %     slowly rotating stars.
% % since we rely on a simple scaling
% % between rotation period and age to make the argument that the rotation period
% % is not caused by incorrect period measurements.
% }
%   \centering
%     \includegraphics[width=1\textwidth]{CMD_cuts}
% \label{fig:CMD_cuts}
% \end{figure}
% To explore the age of this stellar population from a rotation standpoint, it
% was first necessary to remove visual binaries and subgiants from the sample.
% The rotational evolution of these two types of stars is generally different to
% that of single stars which more usually follow a Skumanich-like spin-down law.
% Tidal and magnetic interactions between the two components of a binary system
% can influence the rotation periods of both stars, and the expanding envelopes
% of subgiants drive rapid spin-down through conservation of angular momentum.
We removed visual binaries and subgiants from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
We fit a 6th-order polynomial to the main sequence and raised it by 0.27 dex
to approximate the division between single stars and visual binaries and
removed all stars above this line from the sample.
We also removed subgiants by eliminating stars brighter than 6th magnitude in
\gaia\ G-band.
\begin{figure}
  \caption{
Dereddened MS \kepler\ stars with \mct\ rotation periods on the \gaia\ CMD.
We excluded visual binaries by removing stars above the dashed line.
    Points are colored by their gyrochronal
    age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient_combo}
\label{fig:age_gradient}
\end{figure}

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate stellar velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
We removed stars with negative parallaxes, parallax signal-to-noise ratios
less than 10, stars fainter than 16th magnitude, stars with absolute \vb\
uncertainties greater than 1 \kms\, and stars with galactic latitudes greater
than 15\degrees\ from the sample.
Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
and because latitude is correlated with stellar mass (lower mass stars are
older and tend to be preferentially located at high $b$), we found that
including stars with $b > 15$\degrees\ led to a larger number of \vb\ outliers
at low stellar mass.
15\degrees\ was found to be an adequate compromise between maintaining a close
relationship between \vb\ and \vz\ and keeping a large enough sample size to
provide meaningful results as we found that reducing this to 10\degrees did
not significantly affect our results.
We also found that the anti-correlation between stellar mass and $b$ is
negligable for stars cooler than 5000 K.
For this reason, and because the rotational evolution of old, G stars does not
follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
dwarfs.

>>>>

We used the publicly available \kepler-\gaia\ DR2 crossmatched
catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ catalog of
stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
DR2 catalog of parallaxes, proper motions and apparent magnitudes.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
\citep{green2018}, and {\tt astropy} \citep{astropy2013}.
We estimated effective temperatures from dereddened \Gaia\ \gcolor\ color,
using an 8th-order polynomial relation calibrated using .... stars
\racomment{ask Jason for details}.
\begin{equation}
    \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
    301C^6 - 26C^7,
% 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
%           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
%           301.13205924630165, -25.923997443169355]
\label{eq:curtis}
\end{equation}
where C is \gaia\ \gcolor.

% \begin{figure}
%   \caption{
% A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
%     extinction-corrected magnitudes, colored by rotation period.
% We excluded photometric binaries and subgiants from our analysis by removing
% stars above the two dashed lines.
% % The rotation periods of binaries and subgiants do not follow a Skumanich-like
% % braking law.
% % The rotation period gradient across the main sequence is visible by eye in
% %     this figure: young, rapidly rotating stars are located below the old,
% %     slowly rotating stars.
% % since we rely on a simple scaling
% % between rotation period and age to make the argument that the rotation period
% % is not caused by incorrect period measurements.
% }
%   \centering
%     \includegraphics[width=1\textwidth]{CMD_cuts}
% \label{fig:CMD_cuts}
% \end{figure}
% To explore the age of this stellar population from a rotation standpoint, it
% was first necessary to remove visual binaries and subgiants from the sample.
% The rotational evolution of these two types of stars is generally different to
% that of single stars which more usually follow a Skumanich-like spin-down law.
% Tidal and magnetic interactions between the two components of a binary system
% can influence the rotation periods of both stars, and the expanding envelopes
% of subgiants drive rapid spin-down through conservation of angular momentum.
Visual binaries and subgiants were removed from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and visual binaries.
All stars above this line were removed from the sample.
Subgiants were also removed by eliminating stars brighter than 6th magnitude
in \gaia\ G-band.
\begin{figure}
  \caption{
Dereddened MS \kepler\ stars with \mct\ rotation periods on the \gaia\ CMD.
We excluded visual binaries by removing stars above the dashed line.
    Points are colored by their gyrochronal
    age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient_combo}
\label{fig:age_gradient}
\end{figure}

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate stellar velocities.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
greater than 1 \kms\, and stars with galactic latitudes greater than
15\degrees\ were removed from the sample.
Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
and because latitude is correlated with stellar mass (lower mass stars are
older and tend to be preferentially located at high $b$, although this trends
is greatly reduced for stars cooler than 5000 K), we found that excluding
stars with $b > 15$\degrees\ reduced the number of \vb\ outliers at low
stellar mass.
15\degrees\ was found to be an adequate compromise between maintaining a close
relationship between \vb\ and \vz\ and retaining a sample large enough to
provide meaningful results.
Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
trends in the data.
% We also found that the anti-correlation between stellar mass and $b$ is
% negligable for stars cooler than 5000 K.
% For this reason, and because the rotational evolution of old, G stars does not
% follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
% restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
% dwarfs.

### 2019-08-29 paper/results.tex
some fairly small edits and rewrites. adding some references

To explore the relationship between rotation period, \teff\ and age, we
selected groups of stars within different age ranges \citep[where age was
calculated using the][gyrochronology relation]{angus2019}, and calculated the
velocity dispersion: the standard deviation of velocities, \sigmavb, as a
function of effective temperature for each age group.
We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
and temperature bin to remove non-Gaussian outliers.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars.
Ages were calculated using dereddened \gaia\ \gcolor\ color, however
throughout this paper we show rotation periods as a function of {\it effective
temperature}.
We chose to use effective temperatures and not colors in this analysis as it
is the linear quantity and therefore easier to divide into bins of roughly
equal numbers of stars.

\begin{figure}
  \caption{
Top: rotation period vs effective temperature for stars in the \mct\
    catalog.
    The full catalog, with subgiants and visual binaries removed is shown in
    grey, and stars selected to be in different age groups (between \tmin\ and
    5000 K) are overlayed in color.
These age groups were selected using the \citet{angus2019} gyrochronology
    relation.
The legend in the center of the figure lists the age range, in Gyr, of each
    group.
Bottom: velocity dispersion vs effective temperature for each age
    group.
The color of the line corresponds to the color of the group shown in the top
    panel.
If the gyrochronal model were correct at all ages, and the stars in each group
    were the same age across temperatures, the velocity dispersion would be
    constant as a function of \teff.
However, the velocity dispersions of the oldest age groups increase with
    \teff, indicating the \citet{angus2019} gyrochronology model underpredicts
    the the ages of late-K dwarfs relative to the ages of early K dwarfs at
    old ages.
% An alternative explanation could be that the gyrochronology relation is
%     correct and {\it mass-dependent heating} is responsible for the greater
%     velocity dispersions of cooler stars.
}
  \centering
    \includegraphics[width=1\textwidth]{age_cut}
\label{fig:age_cut}
\end{figure}
The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample
(excluding visual binaries and subgiants) in grey, with coeval groups shown in
color.
The color of the points corresponds to the age ranges specified in the legend
(in Gyr), which also apply to the lines in the lower panel.
The bottom panel shows the velocity dispersion, \sigmavb\ of each age group,
as a function of effective temperature.
% We only included stars within a temperature range of 5000 - 3500 K in our
% analysis, as hotter stars are more likely to have stopped magnetic braking
% \citep{vansaders2016}, which could bias the results.
M dwarfs were not included in our analysis because such faint stars cannot be
observed at large heights above the plane (because of the low galactic
latitude of the \kepler\ field, stars at high-Z are more distant), which
introduces a mass-dependent velocity bias: cooler populations of stars are
skewed towards lower velocity dispersions.
The coolest temperature bins in figure \ref{fig:age_cut} have low velocity
dispersions, indicating that this effect may already become important at
temperatures lower than $\sim$ 4000 K.

Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
with gyrochronal age across all temperatures.
The constant velocity dispersion of young stars as a function of temperature
shows that the Praesepe-calibrated gyrochronology relation accurately predicts
the relative ages of {\it young} field stars.
% To our knowledge, no gyrochronology relation has never been demonstrated to
% correctly predict ages (either relative or absolute) for such cool or such
% young field stars.
% This is not particularly remarkable however, since these young stars are a
% similar age to the Praesepe cluster, which was used to calibrate the
% period-color relation.
% However, if the stars in each selected age group had the {\it same} age across
% the temperature range, their velocity dispersion would be a constant function
% of \teff.
However, although velocity dispersion is constant across temperature for young
stars, it increases as a function of temperature for old stars.
This indicates that the shape of the period-color relation flattens out over
time.

@@

a result of an intrinsically larger velocity dispersion for low-mass stars.
We performed the same analysis on the 537 stars in this sample with RVs in

@@

\begin{figure}
  \caption{
      Top: Stellar velocity (\vb) as a function of \teff\ for
      \kepler\ K and M dwarfs.
Vertical lines indicate different \teff-groupings used to calculate velocity
    dispersion.
Pink stars were not included in velocity dispersion calculations as they were
    either removed as outliers during a sigma clipping process, or they lie at
    the sparcely populated, extremely cool end of the temperature range.
    Velocity dispersion and \teff\ are slightly correlated,
    indicating that mass-dependent heating does not significantly affect
    low-mass stars in the \kepler\ field.
}
  \centering
    \includegraphics[width=1\textwidth]{vb_vs_teff}
\label{fig:vb_vs_teff}
\end{figure}
Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
strongly affect the \mct\ sample of rotating \kepler\ stars.
For this reason, we assume that age difference is the only significant cause
of velocity dispersion differences between groups.
In other words, (\vb) velocity dispersion is a reliable age proxy for the
\mct\ sample.

@@

\begin{figure}
  \caption{
Rotation period vs effective temperature for stars in the \mct\ sample,
    colored by the velocity dispersions of stars in restricted period and
    temperature ranges.
}
  \centering
    \includegraphics[width=1\textwidth]{dispersion_period_teff}
\label{fig:dispersion_period_teff}
\end{figure}
Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
our sample, coloured by (\vb) velocity dispersion.
Here, velocity dispersion was calculated for groups of stars in
uniformly-spaced bins in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating doesn't affect this sample, and \vb\
at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
dispersion can be interpreted as an age proxy and stars plotted in a similar
color in figure \ref{fig:dispersion_period_teff} are similar ages.
Interpreted this way, it appears that rotation period {\it decreases} with
decreasing effective temperature at old ages.  This is a paradigm shift
for gyrochronology because stellar spin-down rate is thought to be directly
tied to magnetic field strength, and the deeper convection zones of cooler
stars generate stronger magnetic fields which {\it should} lead to more
efficient angular momentum loss.
A similar phenomenon has been observed in the 1.1 Gyr NGC 6811 cluster, where
the rotation periods of mid-K dwarfs are faster than expected; their
rotational evolution appears to have stalled, and the period-\teff\ relation
is flat \citep{curtis2019}.
NGC 6811 straddles the rotation period gap and may be the `missing link' that
connects two epoch of stellar spin-down: an early stage where the
period-\teff\ relation for K dwarfs has a negative slope and a late stage
where it has a positive slope.
In this case the period gap may delineate the transition between these two
regimes and is the point at which stellar magnetic dynamos likely undergo a
dramatic structural shift.

>>>>

To explore the relationship between rotation period, \teff\ and age, we
selected groups of stars within different age ranges \citep[where age was
calculated using the][gyrochronology relation]{angus2019}, and calculated the
velocity dispersion: the standard deviation of velocities in the direction of
increasing galactic latitude, \sigmavb, as a function of effective temperature
for each age group.
We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
and temperature bin to remove non-Gaussian outliers.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars.
Ages were calculated using dereddened \gaia\ \gcolor\ color, however
throughout this paper we show rotation periods as a function of effective
temperature, \teff.
We chose to use \teff and not photometric color in this analysis as it is the
linear quantity and therefore easier to divide into bins of roughly equal
numbers of stars.

\begin{figure}
  \caption{
Top: rotation period vs effective temperature for stars in the \mct\
    catalog.
    The full catalog, with subgiants and visual binaries removed is shown in
    grey, and stars selected to be in different age groups (between \tmin\ and
    \tmax\ K) are overlayed in color.
These age groups were selected using the \citet{angus2019} gyrochronology
    relation.
The legend in the center of the figure lists the age range, in Gyr, of each
    group.
Bottom: velocity dispersion vs effective temperature for each age
    group.
The color of the line corresponds to the color of the group shown in the top
    panel.
If the gyrochronal model were correct at all ages, and the stars in each group
    were the same age across temperatures, the velocity dispersion would be
    constant as a function of \teff.
However, the velocity dispersions of the oldest age groups increase with
    \teff, indicating the \citet{angus2019} gyrochronology model underpredicts
    the the ages of late-K dwarfs relative to the ages of early K dwarfs at
    old ages.
% An alternative explanation could be that the gyrochronology relation is
%     correct and {\it mass-dependent heating} is responsible for the greater
%     velocity dispersions of cooler stars.
}
  \centering
    \includegraphics[width=1\textwidth]{age_cut}
\label{fig:age_cut}
\end{figure}
The top panel of figure \ref{fig:age_cut} shows the full \mct\ sample
(excluding visual binaries and subgiants) in grey, with coeval groups shown in
color.
The color of the points corresponds to the age ranges specified in the legend
(in Gyr), which also apply to the lines in the lower panel.
The bottom panel shows the velocity dispersion, \sigmavb\ of each age group,
as a function of effective temperature.
% We only included stars within a temperature range of 5000 - 3500 K in our
% analysis, as hotter stars are more likely to have stopped magnetic braking
% \citep{vansaders2016}, which could bias the results.
Late M dwarfs were not included in our analysis because such faint stars
cannot be observed at large heights above the plane (because of the low
galactic latitude of the \kepler\ field, stars at high-Z are more distant),
which introduces a mass-dependent velocity bias: cooler populations of stars
are skewed towards lower velocity dispersions.
The coolest temperature bins in the lower panel of figure \ref{fig:age_cut}
have low velocity dispersions, indicating that this effect may already become
important at temperatures lower than $\sim$ 4000 K.

Overall, figure \ref{fig:age_cut} shows that velocity dispersion increases
with gyrochronal age across all temperatures, implying that both velocity
dispersion and rotation periods increase with age as expected.
The constant velocity dispersion of young stars as a function of temperature
shows that the Praesepe-calibrated gyrochronology relation accurately predicts
the relative ages of {\it young} field stars.
% To our knowledge, no gyrochronology relation has never been demonstrated to
% correctly predict ages (either relative or absolute) for such cool or such
% young field stars.
% This is not particularly remarkable however, since these young stars are a
% similar age to the Praesepe cluster, which was used to calibrate the
% period-color relation.
% However, if the stars in each selected age group had the {\it same} age across
% the temperature range, their velocity dispersion would be a constant function
% of \teff.
In contrast however, velocity dispersion increases as a function of
temperature for old stars, meaning the \citet{angus2019} gyrochronology
relation underpredicts the ages of old, late-K dwarfs (or overpredicts the
ages of old early-K dwarfs).
This indicates that the relationship between period and photometric
color or \teff\ flattens out over time.
In other words, the lines of constant age sweeping diagonally upwards in the
top panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.

@@

a result of an intrinsically larger velocity dispersion for low-mass stars,
nor is it a result of the selection function.
Since we use \vb, not \vz\ in this analysis, it is possible that the
anti-correlation between \teff\ and galactic latitude, $b$, could influence
these results: stars at higher latitudes would have additional velocity
components in the ${\bf x}$ and ${\bf y}$ directions, which could increase
\vb\ but not \vz.
However, since the relationship between \sigmavb\ and \teff\ is positively,
not negatively correlated for cool stars in the \kepler\ field, this effect is
probably too small to influence our results.
We performed the same analysis on the 537 stars in this sample with RVs using

@@

\begin{figure}
  \caption{
      Top: Stellar velocity (\vb) as a function of \teff\ for
      \kepler\ K and M dwarfs.
Vertical lines indicate different \teff-groupings used to calculate velocity
    dispersion.
Pink stars were not included in velocity dispersion calculations as they were
    either removed as outliers during a sigma clipping process, or they lie at
    the sparcely populated, extremely cool end of the temperature range.
    Velocity dispersion and \teff\ are slightly positively correlated, likely
    due to a brightness-related selection bias, indicating that mass-dependent
    heating does not significantly affect low-mass stars in the \kepler\
    field.
}
  \centering
    \includegraphics[width=1\textwidth]{vb_vs_teff}
\label{fig:vb_vs_teff}
\end{figure}
Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
strongly affect the \mct\ sample of rotating \kepler\ stars.
For this reason, we assume that age difference is the only significant cause
of velocity dispersion differences between groups.
In other words, (\vb) velocity dispersion is a reliable age proxy for the
\mct\ sample.

@@

\begin{figure}
  \caption{
Rotation period vs effective temperature for stars in the \mct\ sample,
    colored by the velocity dispersions of stars in restricted period and
    temperature ranges.
}
  \centering
    \includegraphics[width=1\textwidth]{vplot}
\label{fig:dispersion_period_teff}
\end{figure}
Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
our sample, coloured by (\vb) velocity dispersion.
Here, velocity dispersion was calculated for groups of stars in
uniformly-spaced bins in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating doesn't affect this sample, and \vb\
at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
dispersion can be interpreted as an age proxy and stars plotted in a similar
color in figure \ref{fig:dispersion_period_teff} are similar ages.
Interpreted this way, it appears that rotation period {\it decreases} with
decreasing effective temperature at old ages.  This is a paradigm shift
for gyrochronology because stellar spin-down rate is thought to be directly
tied to magnetic field strength, and the deeper convection zones of cooler
stars generate stronger magnetic fields which {\it should} lead to more
efficient angular momentum loss.
A similar phenomenon has been observed in the 1.1 Gyr NGC 6811 cluster, where
the rotation periods of mid-K dwarfs are faster than expected; their
rotational evolution appears to have stalled, and the period-\teff\ relation
is flat \citep{curtis2019}.
NGC 6811 straddles the rotation period gap and may be the `missing link' that
connects two epoch of stellar spin-down: an early stage where the
period-\teff\ relation for K dwarfs has a negative slope and a late stage
where it has a positive slope.
In this case the period gap may delineate the transition between these two
regimes and is the point at which stellar magnetic dynamos likely undergo a
dramatic structural shift.
